---
title: "R Syntax and Data Structures Answer Key"
author:
  - Will Gammerdinger
date: "2025-05-30"
---

```{r}
#| echo: false
glengths <- c(4.6, 3000, 50000)
species <- c("ecoli", "human", "corn")
number <- 15
```


# Exercise 1

1. Try to create a vector of numeric and character values by *combining* the two vectors that we just created (`glengths` and `species`). Assign this combined vector to a new variable called `combined`. *Hint: you will need to use the combine `c()` function to do this*. 

```{r}
# Assign the vector of glengths and species to combined
combined <- c(glengths, species)
```

2. Print the `combined` vector in the console, what looks different compared to the original vectors?

```{r}
# Print the combined vector to the console
combined
```

`glengths` has been coerced into being a character vector.

# Exercise 2

Let's say that in our experimental analyses, we are working with three different sets of cells: normal, cells knocked out for geneA (a very exciting gene), and cells overexpressing geneA. We have three replicates for each celltype.

1.  Create a vector named `samplegroup` with nine elements: 3 control ("CTL") values, 3 knock-out ("KO") values, and 3 over-expressing ("OE") values.

```{r}
# Create the samplegroup vector with 3 CTL, 3 KO and 3 OE
samplegroup <- c("CTL", "CTL", "CTL", "KO", "KO", "KO", "OE", "OE", "OE")
```

2.  Turn `samplegroup` into a factor data structure.

```{r}
# Convert samplegroup into a factor and re-assign it to samplegroup
samplegroup <- factor(samplegroup)
```

# Exercise 3

Create a data frame called `favorite_books` with the following vectors as columns:

```{r}
# Create a character vector containing book names and assign it to titles
titles <- c("Catch-22", "Pride and Prejudice", "Nineteen Eighty Four")
# Create a numeric vector containing page numbers and assign it to pages
pages <- c(453, 432, 328)
```

```{r}
# Create a data frame of titles and pages
favorite_books <- data.frame(titles, pages)
```

# Exercise 4

Create a list called `list2` containing `species`, `glengths`, and `number`.

```{r}
# Create a list called list2 composed of species, glengths and number
list2 <- list(species, glengths, number)
```
