---
title: "Functions and Arguments Answer Key"
author:
  - Will Gammerdinger
date: "2025-07-01"
---

```{r}
#| label: load_data
#| echo: false
glengths <- c(30, 4.6, 3000, 50000, 90)
```

# Exercise 1

1.  Let's use base R function to calculate **mean** value of the `glengths` vector. You might need to search online to find what function can perform this task.

```{r}
#| label: mean_glengths
# Obtain the mean of glengths
mean(glengths)
```

2.  Create a new vector `test <- c(1, NA, 2, 3, NA, 4)`. Use the same base R function from exercise 1 (with addition of proper argument), and calculate mean value of the `test` vector. The output should be `2.5`.

```{r}
#| label: NA_mean_test
# Create the test vector
test <- c(1, NA, 2, 3, NA, 4)
# Take the mean of the test vector while excluding NA values
mean(test, na.rm = TRUE)
```

3.  Another commonly used base function is `sort()`. Use this function to sort the `glengths` vector in **descending** order.

```{r}
#| label: sort_glengths
# Sort the glengths vector in descending order
sort(glengths, decreasing = TRUE)
```

# Exercise 2

Write a function called `multiply_it`, which takes two inputs: a numeric value `x`, and a numeric value `y`. The function will return the product of these two numeric values, which is `x * y`. For example, `multiply_it(x=4, y=6)` will return output `24`.

```{r}
#| label: create_use_multiply_it_function
# Create the multiply_it function that accepts values of x and y
multiply_it <- function(x,y) {
  # Multiply x and y and assign the product to the object called product
  product <- x * y
  # Return the product object to the console
  return(product)
}

# Using the multiply_it function, multiply x and y using 4 and 6, respectively
multiply_it(x = 4, y = 6)
```

