---
title: "Reading in and inspecting data Answer Key"
author:
  - Will Gammerdinger
date: "2025-07-01"
---

```{r}
#| label: load_data
#| echo: false
# Read in metadata
metadata <- read.csv(file="data/mouse_exp_design.csv")

# Load glengths
glengths <- c(30, 4.6, 3000, 50000, 90)

# Create samplegroup vector
samplegroup <- c("CTL", "CTL", "CTL", "KO", "KO", "KO", "OE", "OE", "OE")
# Turn samplegroup into a factor
samplegroup <- factor(samplegroup)
```

# Exercise 1

1. Inside your project's `data` folder you should see a file called `project-summary.txt`. Read it in to R using `read.table()` with the appropriate arguments and store it as the variable `proj_summary`. To figure out the appropriate arguments to use with `read.table()`, keep the following in mind:
	- all the columns in the input text file have column name/headers
	- you want the first column of the text file to be used as row names (hint: look up the input for the `row.names =` argument in `read.table()`)

```{r}
#| label: read_proj_summary
# Read in project-summary.txt from the data directory and let R know there is a header and that you want the first column to be converted to the rownames then assign it to an object called proj_summary
proj_summary <- read.table("data/project-summary.txt", header = TRUE, row.names = 1)
```

2. Display the contents of `proj_summary` in your console

```{r}
#| label: view_proj_summary
# Display the contents of proj_summary
proj_summary
```

# Exercise 2

1. Use the `class()` function on `glengths` and `metadata`, how does the output differ between the two?

```{r}
#| label: view_classes
# Return the class of glengths
class(glengths)
# Return the class of metadata
class(metadata)
```

`glengths` is a numeric vector and `metadata` is a data frame.

2. Use the `summary()` function on the `proj_summary` dataframe, what is the median "rRNA_rate"?

```{r}
#| label: summary_proj_summary
# Provide a summary of the proj_summary object
summary(proj_summary)
```

The median "rRNA_rate" is 0.005345.

3. How long is the `samplegroup` factor?

```{r}
#| label: length_samplegroup
# Return the length of the samplegroup factor vector
length(samplegroup)
```

4. What are the dimensions of the `proj_summary` dataframe?

```{r}
#| label: dimensions_proj_summary
# Return the dimensions of the proj_summary dataframe
dim(proj_summary)
```

5. When you use the `rownames()` function on `metadata`, what is the *data structure* of the output?

```{r}
#| label: structure_metadata_rownames
# Return the data structure for the output of rownames() on the metadata data frame object
str(rownames(metadata))
```

It is a character vector.

[Optional] How many elements in (how long is) the output of `colnames(proj_summary)`? Don't count, but use another function to determine this.

```{r}
#| label: number_columns_proj_summary
# Return the number of elements in colnames(proj_summary)
length(colnames(proj_summary))
```
