---
title: "Practicing with Custom Functions Answer Key"
author:
  - Mary Piper
  - Meeta Mistry
  - Radhika Khetani
  - Will Gammerdinger
date: "2025-09-05"
---

## Exercises

1. **Custom Functions** - Let's create a function `temp_conv()`, which converts the temperature in Fahrenheit (input) to the temperature in Kelvin (output). _Let's perform a two-step calculation: first convert from Fahrenheit to Celsius, and then convert from Celsius to Kelvin._

    * The formula for celsius to farenheight: `temp_c = (temp_f - 32) * 5 / 9`
    * The formula for celsius to kelvin: `temp_k = temp_c + 273.15`
    * Test your function. If your input is 70, the result of `temp_conv(70)` should be 294.2611.
    
```{r}
#| label: create_temp_conv_function
# Create temp_conv_function
temp_conv <- function(temp_f) {
  temp_c = (temp_f - 32) * 5 / 9
  temp_k = temp_c + 273.15
  return (temp_k)
}
```

2. **Nesting Functions** - Now we want to round the temperature in Kelvin (output of `temp_conv()`) to a single decimal place. 
     
     * Use the `round()` function with the newly-created  `temp_conv()` function to achieve this in one line of code. 
     * If your input is 70, the output should now be 294.3.

```{r}
#| label: nested_round_function
# Round output of temp_conv to 1 digit
round(temp_conv(70), digits = 1)
```