---
title: "Custom functions for consistent plots"
author: "Mary Piper, Meeta Mistry, Radhika Khetani, Will Gammerdinger"
date: "Wednesday, February 4, 2021"
---

Approximate time: 20 minutes

```{r}
#| echo: false
#| eval: true
library(ggplot2)
load("data/new_metadata.RData")
```

## Learning Objectives 

* Apply the custom function to generate consistent plots.

## Consistent formatting using custom functions

When publishing, it is helpful to ensure all plots have similar formatting. To do this we can create a custom function with our preferences for the theme. Remember the structure of a function is:

```{r}
#| eval: false
name_of_function <- function(arguments) {
    statements or code that does something
}
```

Now, let's suppose we always wanted our theme to include the following:

```{r}
#| eval: false
theme_bw() +
theme(axis.title=element_text(size=rel(1.5))) +
theme(plot.title=element_text(size=rel(1.5), hjust=0.5))
```

::: callout-note
You can also combine multiple arguments within the same theme() function:

```{r}
#| eval: false
theme_bw() +
theme(axis.title=element_text(size=rel(1.5)), plot.title=element_text(size=rel(1.5), hjust=0.5))
```
:::

If there is nothing that we want to change when we run this, then we do not need to specify any arguments. Creating the function is simple; we can just put the code inside the `{}`:

```{r}
personal_theme <- function(){
  theme_bw() +
  theme(axis.title=element_text(size=rel(1.5))) +
  theme(plot.title=element_text(size=rel(1.5), hjust=0.5))
}
```

Now to run our personal theme with any plot, we can use this function in place of the lines of `theme()` code:

```{r}
ggplot(new_metadata) +
  geom_point(aes(x=age_in_days, y=samplemeans, color=genotype, shape=celltype), size=rel(3.0)) +
  xlab("Age (days)") +
  ylab("Mean expression") +
  ggtitle("Expression with Age") +
  personal_theme()
```
