---
title: "Plotting with the ggplot2 Package Answer Key"
author:
  - Will Gammerdinger
date: "2025-07-01"
---

```{r}
#| label: load_libraries_data
#| echo: false
# Load libraries
library(ggplot2)

# Load data
load("data/new_metadata.RData")
```

# Exercise 1

1. The current axis label text defaults to what we gave as input to `geom_point` (i.e the column headers). We can change this by **adding additional layers** called `xlab()` and `ylab()` for the x- and y-axis, respectively. Add these layers to the current plot such that the x-axis is labeled "Age (days)" and the y-axis is labeled "Mean expression".

```{r}
#| label: added_axes_labels
# Add axes labels to the plot
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression")
```

2. Use the `ggtitle` layer to add a plot title of your choice. 

```{r}
#| label: add ggtitle
# Add title to the plot
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression") +
  ggtitle("Mean Expression by Age")
```

3. Add the following new layer to the code chunk `theme(plot.title=element_text(hjust=0.5))`.

```{r}
#| label: center_plot_title
# Center plot title
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression") +
  ggtitle("Mean Expression by Age") +
  theme(plot.title=element_text(hjust=0.5))
```

* What does it change?

  It centers the title   
   
* How many theme() layers can be added to a ggplot code chunk, in your estimation?

  There is no limit to the number of theme() layers that can be added to a ggplot code chunk.